Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-01-06 paper/ms.tex
Corrected definition of \eta.
figure path
Fixed paths to figure in ms.tex
tiny typographical change
proposal: re-define S_0
comment on definition of S_0
reverting S_0 definition
Started writing discussion section.

These identities yield a kernel of the form
\begin{equation}
k_{SHO}(\tau) = S_0 \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \begin{cases}
\cosh{(\eta \omega_0 \tau)} + (2Q\eta)^{-1} \sinh{(\eta \omega_0 \tau)}, & 0 < Q < 1/2\\
2 (1+\omega_0 \tau), & Q = 1/2\\
\cos{(\eta \omega_0 \tau)} + (2Q\eta)^{-1} \sin{(\eta \omega_0 \tau)},& 1/2 < Q < \infty\\
\end{cases}
\end{equation}
where $\eta = \vert 1-(4Q^2)^{-1}\vert$.
It is interesting to note that, because of the damping, the characteristic
oscillation frequency in this model $d_j$, for any finite quality factor $Q > 1/2$,
is not equal to the frequency of the undamped oscillator $\omega_0$.

@@

{\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
\begin{eqnarray}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
\end{eqnarray}
A model like this has been used to model the granulation is Solar data
\citep{Michel:2009}}

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/sho.pdf}
\caption{(left) The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $c_j = 2\,Q = 20$ and normalized so that
    $S(d_j) = Q^2 = 100$.
    (right) The corresponding autocorrelation functions with the same colors.
    \figurelabel{sho}}
\end{center}
\end{figure}

@@

@@

For the complex values with $d_j \ne 0$, the equations are modified as follows.  We introduce complex
auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
r_k^R - i r_k^I$, with $k \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric
set of equations, and since the values of these variables are discarded in the end, their sign
is irrelevant).  The length of the vectors $l_k$ and $r_k$
is the number of kernel components, $J$;  we denote the individual elements of these vectors
as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,J\}$.  Note that the first $J_0$ elements we take
to be the kernels with $d_j=0$, and for these we do not include the imaginary component
equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
components as well:  $\gamma_k = \left[\exp{(-(c_1+id_1) t_{k,k+1})} ... \exp{(-(c_J+id_J) t_{k,k+1})}\right]^T
 = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-c_j t_{k,k+1}}
\cos{(d_j t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-c_j t_{k,k+1}}\sin{(d_j t_{k,k+1})}$.
Equation (61) becomes:
\begin{equation} \label{Amb61}
% See 9/2/2016 notes
\sum_{j=1}^J \left(\frac{a_j}{2} l_{k,j}^R +\frac{b_j}{2} l_{k,j}^I\right)+ \frac{d}{2} x_k + \sum_{j=1}^J \left[
\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{h_k}{2},
\end{equation}
for $j \in \{1,...,J\}$ and $k \in \{1,...,N\}$, where $h_j$ is the $j$th data point.  There is no imaginary component to equation (61)
in \citet{Ambikasaran:2015}.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.175]{./twoterm.pdf}
\hspace{0.2in}
\includegraphics[scale=0.85]{./colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix where $N=10$, $J_0=0$, and $J=2$, following
the example shown in \citet{Ambikasaran:2015}.}
\figurelabel{matrix_structure}
\end{figure}

>>>>

These identities yield a kernel of the form
\begin{equation}
k(\tau) = S_0 \omega_0 Q e^{-\frac{\omega_0 \tau}{2Q}} \begin{cases}
\cosh{(\eta \omega_0 \tau)} + \frac{1}{2\eta Q} \sinh{(\eta \omega_0 \tau)}, & 0 < Q < 1/2\\
2 (1+\omega_0 \tau), & Q = 1/2\\
\cos{(\eta \omega_0 \tau)} + \frac{1}{2\eta Q} \sin{(\eta \omega_0 \tau)},& 1/2 < Q < \infty\\
\end{cases}
\end{equation}
where $\eta = \vert 1-(4Q^2)^{-1}\vert^{1/2}$.
It is interesting to note that, because of the damping, the characteristic
oscillation frequency in this model $d_j$, for any finite quality factor $Q > 1/2$,
is not equal to the frequency of the undamped oscillator $\omega_0$.

@@

{\item For $Q = 1/\sqrt{2}$, \eq{sho-psd} simplifies to
\begin{eqnarray}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0}{(\omega/\omega_0)^4+1} \quad.
\end{eqnarray}
A model like this has been used to model the granulation is Solar data
\citep{Michel:2009}}, which has a kernel of
\begin{equation}
k(\tau) = S_0 \omega_0e^{-\frac{1}{\sqrt{2}}\omega_0\tau} \cos{\left(\frac{\omega_0\tau}{\sqrt{2}}-\frac{\pi}{4}\right)}.
\end{equation}

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/sho.pdf}
\caption{(left) The power spectrum of a stochastically-driven simple harmonic
    oscillator (\eqalt{sho-psd}) plotted for several values of the quality
    factor $Q$.
    For comparison, the dashed line shows the Lorentzian function from
    \eq{lorentz-psd} with $\omega_0/c_j = 2\,Q = 20$ and normalized so that
    $S(d_j)/S(0) = Q^2 = 100$.
    (right) The corresponding autocorrelation functions with the same colors.
    \figurelabel{sho}}
\end{center}
\end{figure}

@@

\section{Discussion}

There are two primary applications we imagine for the \genrp formalism: 1). modeling
astrophysical variability to infer the properties of the object that is varying; 2).
account for astrophysical variability as a source of noise when trying to detect 
additional phenomena; 3). interpolate or extrapolate variations to future or missing
times.  Some examples of the first are determining the characteristic
variability timescale of quasars, measuring the asteroseismic variations of stars,
or detection of quasi-periodic variability in a high-energy source.  Examples of the
second are detecting transiting exoplanets and correcting for stellar activity in
radial velocity measurements. Examples of the third are measuring time delays in 
multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.

Our background is in studying transiting exoplanets, which has only recently begun
to adopt full covariance matrices in analyzing the noise in transiting planet lightcurves.
One early successful attempt at accounting for correlated noise was the wavelet approach
of Carter \& Winn (2010).  This approach allowed for a power-law spectrum for the
noise scaling as $\omega^{-\alpha}$, which unfortunately is not a normalizable, nor
physically-accurate description of stellar variability.  Their approach runs in
$\cal{O}(N)$ time for $\alpha=1$, which is `flicker' or pink noise;  however, this
particular type of noise overpredicts the power spectrum at low frequency compared
to the granulation power spectrum.  It also cannot describe quasi-periodic noise.

@@

% Reprint Ambikasaran equations (59)-(61)

For the complex values with $d_j \ne 0$, the equations are modified as follows.  We introduce complex
auxiliary vectors $l_k = l_{k}^R - i l_{k}^I$, with $k \in \{1,2,...,N\}$ and $l_1 = 0$; $r_k =
r_k^R - i r_k^I$, with $k \in \{2,...,N\}$ and $r_{N+1} = 0$.  Note that we have defined the
imaginary component as being negative for aesthetic reasons:  this yields a symmetric extended
matrix (unfortunately the extended matrix has negative eigenvalues, making it non positive-definite,
so that Cholesky decomposition cannot be applied; nevertheless we prefer to use a symmetric
set of equations, and since the values of these variables are discarded in the end, their sign
is irrelevant).  The length of the vectors $l_k$ and $r_k$
is the number of kernel components, $J$;  we denote the individual elements of these vectors
as $l_{k,j}$ and $r_{k,j}$, $j \in \{1,...,J\}$.  Note that the first $J_0$ elements we take
to be the kernels with $d_j=0$, and for these we do not include the imaginary component
equations, which are unnecessary and only add extra computation time.  We allow each vector $\gamma_k$ to have real and imaginary
components as well:  $\gamma_k = \left[\exp{(-(c_1+id_1) t_{k,k+1})} ... \exp{(-(c_J+id_J) t_{k,k+1})}\right]^T
 = \gamma_k^R + i \gamma_k^I.$  These elements are given by $\gamma_{k,j}^R = e^{-c_j t_{k,k+1}}
\cos{(d_j t_{k,k+1})}$ and $\gamma_{k,j}^I = -e^{-c_j t_{k,k+1}}\sin{(d_j t_{k,k+1})}$.
Equation (61) becomes:
\begin{equation} \label{Amb61}
% See 9/2/2016 notes
% Use n rather than k for {1...N} - use g, h & u, v?
% \gamma & \phi - real & imaginary 
\sum_{j=1}^J \left(\frac{a_j}{2} l_{k,j}^R +\frac{b_j}{2} l_{k,j}^I\right)+ \frac{d}{2} x_k + \sum_{j=1}^J \left[
\gamma_{k,j}^R r_{k+1,j}^R + \gamma_{k,j}^I r_{k+1,j}^I\right] = \frac{y_k}{2},
\end{equation}
for $j \in \{1,...,J\}$ and $k \in \{1,...,N\}$, where $h_j$ is the $j$th data point.  There is no imaginary component to equation (61)
in \citet{Ambikasaran:2015}.

@@

\begin{figure}[!htbp]
\begin{center}
\includegraphics[scale=0.175]{figures/twoterm.pdf}
\hspace{0.2in}
\includegraphics[scale=0.85]{figures/colorcode.pdf}
\end{center}
\caption{Pictorial description of the extended sparse matrix where $N=10$, $J_0=0$, and $J=2$, following
the example shown in \citet{Ambikasaran:2015}.}
\figurelabel{matrix_structure}
\end{figure}
